---
title: "FA6"
author: "Espiritu, Joseph Raphael, Harneyyer, Clores"
date: "2026-05-01"
output:
  pdf_document: default
  html_document: default
---

```{r echo=TRUE}

# Install if needed
# install.packages(c("tidyverse","tidymodels","nnet","GGally"))

library(tidyverse)
library(tidymodels)
library(janitor)

# Load dataset
df <- read_csv("customer_segmentation.csv")
df <- df %>%
  clean_names()
# Preview
colnames(df)
glimpse(df)
str(df)

```

### Data Exploration:

```{r echo=TRUE}

#Missing Values
df %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing_Count")

# Target Dist

df %>%
  count(customer_segment) %>%
  ggplot(aes(x = customer_segment, y = n, fill = customer_segment)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.5) +
  coord_cartesian(ylim = c(3300, 3700)) +
  labs(
    title = "Customer Segment Distribution",
    x = "Customer Segment",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Dist
df %>%
  pivot_longer(c(age, annual_income_k, average_spend_per_visit)) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "white") +
  facet_wrap(~name, scales = "free") +
  labs(
    title = "Distribution of Key Continuous Variables",
    x = "Value",
    y = "Frequency"
  ) +
  theme_minimal()

ggplot(df, aes(x = annual_income_k, fill = customer_segment)) +
  geom_density(alpha = 0.4)

df %>%
  group_by(customer_segment) %>%
  summarise(
    mean_income = mean(annual_income_k),
    mean_spend = mean(average_spend_per_visit)
  )

cor(df %>% select(age, annual_income_k, average_spend_per_visit))

```

### Data Preprocessing:

```{r echo=TRUE}

# Convert categorical variables into factors so R treats them as categories
df <- df %>%
  mutate(
    customer_segment = factor(customer_segment),
    gender = factor(gender),
    product_category_purchased = factor(product_category_purchased)
  )

# Display structure of the dataset to verify variable types
str(df)

# Set seed for reproducibility so splits are consistent across runs
set.seed(123)


# Split data into 80% training and 20% testing while preserving class distribution
split <- initial_split(df, prop = 0.8, strata = customer_segment)

# Extract training dataset
train <- training(split)

# Extract testing dataset
test  <- testing(split)

# Check total number of rows in original dataset
nrow(df)

# Check number of rows in training set
nrow(train)

# Check number of rows in testing set
nrow(test)

# Verify class distribution in training set
train %>% count(customer_segment)

# Verify class distribution in testing set
test %>% count(customer_segment)

rec <- recipe(customer_segment ~ ., data = train) %>%
  update_role(customer_id, new_role = "id") %>%
  step_dummy(all_nominal_predictors()) %>%
  step_normalize(all_numeric_predictors())

# Create preprocessing recipe defining transformations for modeling
rec <- recipe(customer_segment ~ ., data = train) %>%
  update_role(customer_id, new_role = "id") %>%   # Mark ID column so it is not used as a predictor
  step_dummy(all_nominal_predictors()) %>%        # Convert categorical predictors into dummy variables
  step_normalize(all_numeric_predictors())        # Scale numeric predictors to mean 0 and standard deviation 1

# Print recipe structure to review preprocessing steps
rec

# Summarize variables, roles, and types in the recipe
summary(rec)

# Prepare the recipe by estimating required parameters (means, SDs, levels)
prep_rec <- prep(rec)

# Apply preprocessing transformations to training data
train_processed <- bake(prep_rec, new_data = train)

# Apply same transformations to testing data
test_processed  <- bake(prep_rec, new_data = test)

# Remove customer_id column since it should not be used in modeling
train_processed <- train_processed %>%
  select(-customer_id)

# Remove customer_id from testing data as well
test_processed <- test_processed %>%
  select(-customer_id)

# Inspect structure of processed training data to confirm transformations
glimpse(train_processed)

# Check if customer_id column was successfully removed
"customer_id" %in% colnames(train_processed)

# Check summary statistics of scaled age variable (should be centered around 0)
summary(train_processed$age)

# View column names to confirm dummy variables and processed features
colnames(train_processed)

# Check for any remaining missing values in processed data
colSums(is.na(train_processed))

# Confirm target variable remains a factor after preprocessing
class(train_processed$customer_segment)
```

### Model Building:

```{r echo=TRUE}
# Define a multinomial logistic regression model with no regularization (baseline model)
model <- multinom_reg(penalty = 0, mixture = 0) %>%
  set_engine("nnet") %>%                     # Use nnet as the underlying algorithm
  set_mode("classification")                # Specify classification task

# Create a workflow combining preprocessing (recipe) and the model
wf <- workflow() %>%
  add_recipe(rec) %>%                       # Attach preprocessing steps
  add_model(model)                          # Attach model

# View workflow structure
wf

# Fit the model using training data (recipe is applied internally)
fit_model <- fit(wf, data = train)

# View fitted model details
fit_model

# Generate predicted class labels for test data
pred_class <- predict(fit_model, test)

# Generate predicted probabilities for each class
pred_prob <- predict(fit_model, test, type = "prob")

# Combine test data, predicted class, and probabilities into one dataset
pred <- bind_cols(test, pred_class, pred_prob)

# View final prediction dataset
pred

# View column names for the predcition dataset
colnames(pred)
```

### Model Evaluation:

```{r echo=TRUE}

# Compute overall performance metrics such as accuracy, precision, recall, and F1-score
metrics(pred, truth = customer_segment, estimate = .pred_class)

# Create a confusion matrix comparing predicted classes to actual classes
cm <- conf_mat(pred, truth = customer_segment, estimate = .pred_class)

# Display the confusion matrix in table form to observe correct and incorrect classifications
cm

# Visualize the confusion matrix as a heatmap for easier interpretation of classification patterns
autoplot(cm, type = "heatmap")

# Recompute performance metrics (useful if you want to explicitly display results again)
metrics(pred, truth = customer_segment, estimate = .pred_class)

# Compute multinomial log-loss using predicted probabilities for each class
mn_log_loss(
  pred,
  truth = customer_segment,
  `.pred_Budget Shopper`,     # Probability of Budget Shopper
  `.pred_Premium Shopper`,    # Probability of Premium Shopper
  `.pred_Regular Shopper`     # Probability of Regular Shopper
)

# Display detailed model summary including coefficients and standard errors
summary(fit_model$fit$fit)

# Extract model coefficients for each class to understand feature influence
coef(fit_model$fit$fit)
```

### Refinement:

```{r echo=TRUE}
# Create a new preprocessing recipe including an interaction term between income and age
rec2 <- recipe(customer_segment ~ ., data = train) %>%
  update_role(customer_id, new_role = "id") %>%          # Mark customer_id as an identifier so it is excluded from modeling
  step_interact(~ annual_income_k:age) %>%               # Create interaction feature capturing combined effect of income and age
  step_dummy(all_nominal_predictors()) %>%               # Convert categorical predictors into numeric dummy variables
  step_normalize(all_numeric_predictors())               # Standardize numeric predictors (mean = 0, sd = 1)

# Define a multinomial logistic regression model with tunable regularization strength (penalty)
model_tuned <- multinom_reg(penalty = tune(), mixture = 0) %>%
  set_engine("nnet") %>%                                # Use nnet implementation for multinomial regression
  set_mode("classification")                            # Specify classification task

# Combine preprocessing (rec2) and model into a single workflow pipeline
wf2 <- workflow() %>%
  add_recipe(rec2) %>%                                  # Attach updated preprocessing steps
  add_model(model_tuned)                                # Attach tunable model

# Set seed to ensure reproducibility of cross-validation results
set.seed(123)

# Create 5-fold cross-validation splits while preserving class distribution
folds <- vfold_cv(train, v = 5, strata = customer_segment)

# Define a grid of penalty values to test (log scale from 10^-3 to 10^1)
grid <- grid_regular(penalty(range = c(-3, 1)), levels = 5)

# Perform hyperparameter tuning across folds using the defined grid and evaluate using accuracy
tuned <- tune_grid(
  wf2,
  resamples = folds,                                    # Apply cross-validation
  grid = grid,                                          # Test multiple penalty values
  metrics = metric_set(accuracy)                        # Evaluate using classification accuracy
)

# Select the best-performing penalty value based on highest accuracy
best_params <- select_best(tuned, metric = "accuracy")

# View the optimal hyperparameter value selected from tuning
best_params

# Finalize workflow by inserting the best penalty value into the model
final_wf <- finalize_workflow(wf2, best_params)

# Fit the final optimized model using the full training dataset
final_fit <- fit(final_wf, data = train)

# Generate predictions and probabilities on the test dataset and combine into one dataset
pred <- bind_cols(
  test,
  predict(final_fit, test),                              # Predicted class labels
  predict(final_fit, test, type = "prob")                # Predicted class probabilities
)

# Evaluate final model accuracy by comparing predicted vs actual classes
metrics(pred, truth = customer_segment, estimate = .pred_class)

# Generate confusion matrix to assess classification performance across all classes
cm2 <- conf_mat(pred, truth = customer_segment, estimate = .pred_class)

# Visualize the confusion matrix as a heatmap for easier interpretation of classification patterns
autoplot(cm2, type = "heatmap")
```

### Reportings:

#### 1. Introduction

  This analysis aims to predict customer segments (Budget, Regular, Premium) using demographic and behavioral features such as age, income, spending, and visit frequency. A multinomial logistic regression model was applied, followed by refinement and evaluation to assess predictive performance.
  
#### 2. Data Exploration  
  Observations: ~10,000+ customers  
  Target variable: customer_segment (3 classes)  
  Predictors:  
    Age  
    Annual Income  
    Average Spend per Visit  
    Number of Visits  
    Gender  
    Product Category Purchased  
    
```{r echo=FALSE}
str(df)

df %>%
  count(customer_segment) %>%
  ggplot(aes(x = customer_segment, y = n, fill = customer_segment)) +
  geom_col() +
  geom_text(aes(label = n), vjust = -0.5) +
  coord_cartesian(ylim = c(3300, 3700)) +
  labs(
    title = "Customer Segment Distribution",
    x = "Customer Segment",
    y = "Count"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

Interpretation:  
  Classes are balanced  
  No class imbalance → model should not be biased  
  
```{r echo=FALSE}
df %>%
  pivot_longer(c(age, annual_income_k, average_spend_per_visit)) %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 25, fill = "steelblue", color = "white") +
  facet_wrap(~name, scales = "free") +
  labs(
    title = "Distribution of Key Continuous Variables",
    x = "Value",
    y = "Frequency"
  ) +
  theme_minimal()
```

Interpretation:  
  Variables are reasonably distributed  
  No extreme skewness  
  Suitable for modeling after scaling  
  
```{r echo=FALSE}
ggplot(df, aes(x = annual_income_k, fill = customer_segment)) +
  geom_density(alpha = 0.4)

df %>%
  group_by(customer_segment) %>%
  summarise(
    mean_income = mean(annual_income_k),
    mean_spend = mean(average_spend_per_visit)
  )

cor(df %>% select(age, annual_income_k, average_spend_per_visit))
```

### Important Observation:  
  Heavy overlap between all segments  
  No visible separation  
  Features do not differentiate customer groups  
  Means are nearly identical across groups  
  No meaningful differences  
  Very weak relationships between variables  
  Low predictive structure  

#### 3. Data Preprocessing

  Prior to model development, the dataset underwent several preprocessing steps to ensure that variables were in an appropriate format for modeling and that the data was clean, consistent, and suitable for machine learning algorithms.  
  
  - **Handling Categorical Variables**: Categorical predictors  
  This allows the modeling pipeline to correctly apply transformations such as dummy encoding.  
  
Train-Test Split  

Training set (80%): Used to train the model  
Testing set (20%): Used to evaluate model performance  

Stratified sampling was applied based on customer_segment to preserve the original class distribution in both subsets.

A random seed (set.seed(123)) was set to ensure reproducibility of the split.

```{r echo=FALSE}
# Check summary statistics of scaled age variable (should be centered around 0)
summary(train_processed$age)

# View column names to confirm dummy variables and processed features
colnames(train_processed)

# Check for any remaining missing values in processed data
colSums(is.na(train_processed))

# Confirm target variable remains a factor after preprocessing
class(train_processed$customer_segment)
```

#### Data Validation Checks  

  No missing values were present after preprocessing  
  All predictors were numeric and properly scaled  
  The target variable (customer_segment) remained a factor  
  The dataset structure was consistent and model-ready
  
#### 4. Model Building and Evaluation

  A multinomial logistic regression model was developed to predict customer segment membership (Budget, Regular, Premium). The model was implemented using the tidymodels framework with preprocessing steps applied through a workflow to ensure consistency and prevent data leakage.
  
```{r echo=FALSE}
model <- multinom_reg(penalty = 0, mixture = 0) %>%
  set_engine("nnet") %>%
  set_mode("classification")

wf <- workflow() %>%
  add_recipe(rec) %>%
  add_model(model)

fit_model <- fit(wf, data = train)

pred <- bind_cols(
  test,
  predict(fit_model, test),
  predict(fit_model, test, type = "prob")
)

metrics(pred, truth = customer_segment, estimate = .pred_class)
```

  The multinomial logistic regression model achieved an overall accuracy of 34.5%. Given that the dataset contains three equally distributed classes, this performance is only marginally higher than random chance (approximately 33%).

  The Kappa statistic was 0.0173, indicating that the model provides negligible improvement over chance-level classification. This suggests that the model has very limited predictive capability.
  
```{r echo=FALSE}
cm <- conf_mat(pred, truth = customer_segment, estimate = .pred_class)
autoplot(cm, type = "heatmap")
```

  The confusion matrix reveals substantial misclassification across all customer segments. While some correct classifications are observed along the diagonal, a large number of observations are incorrectly assigned to other classes.

  Notably, each predicted class contains a similar distribution of true classes, indicating that the model fails to effectively distinguish between customer segments. This pattern suggests a lack of separability in the feature space.
  
```{r echo=FALSE}
mn_log_loss(
  pred,
  truth = customer_segment,
  `.pred_Budget Shopper`,
  `.pred_Premium Shopper`,
  `.pred_Regular Shopper`
)
```

  The log-loss value of approximately 1.10 indicates poor probabilistic performance. This value is comparable to that of a random classifier, suggesting that the predicted probabilities are not well-calibrated and do not meaningfully reflect class membership.
  
#### Overall Interpretation of Model

  The multinomial logistic regression model demonstrates poor performance across all evaluation metrics. The low accuracy and near-zero Kappa value indicate that the model is unable to capture meaningful relationships between the predictor variables and the target variable.

  The confusion matrix further confirms that predictions are widely distributed across classes, with no clear pattern of correct classification. This indicates that the model is unable to form decision boundaries that effectively separate customer segments.
  
#### 5. Refinement Strategy and Analysis

To improve model performance, two primary modifications were introduced:  

Feature Interaction  
An interaction term between annual income and age was added to capture potential combined effects on customer segmentation.  
Hyperparameter Tuning  
The regularization parameter (penalty) of the multinomial logistic regression model was tuned using 5-fold cross-validation to optimize model generalization and reduce overfitting.  

```{r echo=FALSE}
rec2 <- recipe(customer_segment ~ ., data = train) %>%
  update_role(customer_id, new_role = "id") %>%
  step_interact(~ annual_income_k:age) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_normalize(all_numeric_predictors())

model_tuned <- multinom_reg(penalty = tune(), mixture = 0) %>%
  set_engine("nnet") %>%
  set_mode("classification")

wf2 <- workflow() %>%
  add_recipe(rec2) %>%
  add_model(model_tuned)

set.seed(123)
folds <- vfold_cv(train, v = 5, strata = customer_segment)

grid <- grid_regular(penalty(range = c(-3, 1)), levels = 5)

tuned <- tune_grid(
  wf2,
  resamples = folds,
  grid = grid,
  metrics = metric_set(accuracy)
)

best_params <- select_best(tuned, metric = "accuracy")

final_wf <- finalize_workflow(wf2, best_params)

final_fit <- fit(final_wf, data = train)

pred <- bind_cols(
  test,
  predict(final_fit, test),
  predict(final_fit, test, type = "prob")
)

metrics(pred, truth = customer_segment, estimate = .pred_class)
```

  The refined multinomial logistic regression model achieved an accuracy of 33.4%, which is slightly lower than the baseline model (34.5%). The Kappa statistic decreased to approximately 0.001, indicating no meaningful improvement over chance-level classification.

  These results suggest that the introduced interaction term and hyperparameter tuning did not enhance the model’s ability to correctly classify customer segments.
  
```{r echo=FALSE}
# Generate confusion matrix to assess classification performance across all classes
cm2 <- conf_mat(pred, truth = customer_segment, estimate = .pred_class)

# Visualize the confusion matrix as a heatmap for easier interpretation of classification patterns
autoplot(cm2, type = "heatmap")
```

  The confusion matrix for the refined model shows a similar pattern to the baseline model, with substantial misclassification across all customer segments. There is no significant improvement in the concentration of correct predictions along the diagonal, indicating that the model continues to struggle in distinguishing between classes.
  
### Analysis:
  The lack of improvement after refinement indicates that the limitation lies not in model complexity or parameter tuning, but in the underlying data structure. Despite introducing interaction effects and optimizing regularization, the model remains unable to identify meaningful decision boundaries between customer segments.

  The optimal penalty value selected during tuning was relatively high, suggesting that strong regularization was necessary to prevent overfitting. However, this also reduced the model’s flexibility, further limiting its predictive capability.
